Efficiently summarizing relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-20 treestats_paper.tex
spell check with en_US.

Genotype matrices are massively redundant, however, and several specialised
compression methods have been proposed~\citep{
christley2008human,qiao2012handling,durbin2014efficient,sambo2014compression,layer2016efficient,
danek2018gtc,lin2019sparse}.
The fundamental reason for this redundancy is that samples share ancestry:
related individuals will tend to share the same state at a particular variant site.
Trees describe these ancestral
relationships~\citep{felsenstein2004inferring,semple2003phylogenetics},
and provide a natural and elegant way of encoding genetic data,
not only recording the history of a sample but also massively compressing
the genotype matrix~\citep{ane2005missing}. Until recently, however,
tree based approaches could not be used to encode data from sexually
reproducing organisms as recombination results in many trees along the
genome~\citep{griffiths1991two,griffiths1996ancestral,rasmussen2014genome},
and there was no efficient way of representing this information.
The \emph{succinct tree sequence} (or \emph{tree sequence}, for brevity)
is a recently introduced data structure that
encodes these genealogical trees along a genome concisely.
It was introduced in the context of coalescent
simulation~\citep{kelleher2016efficient}, leading to scalability increases of
several orders of magnitude over existing methods. The methods were subsequently extended
and refined for forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree}, with similarly large efficiency gains.
Recent work has shown that tree sequence algorithms can
also be used to massively increase the scalability of methods for inferring
genome-wide genealogies, and making it possible to infer trees for millions of
samples~\citep{kelleher2019inferring}.
The key to the remarkable efficiency of tree sequence
algorithms is the way that shared structure in adjacent trees along the genome is encoded.
As one looks across the genome, genealogies change at recombination breakpoints in ancestors.
However, nearby trees tend to share much of their structure,
and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
are often shared across relatively long distances,
which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
This redundancy can be exploited to store genome sequence and the
associated genealogical relationships very compactly, and has formed the basis for
efficient algorithms to calculate several population genetics
statistics~\citep{kelleher2016efficient}.
This paper generalizes the approach to a much broader class of statistics,
supporting arbitrary tree topologies and patterns of allelic variation.
% Want to emphasise generality here: people might assume we're only dealing
% with the biallelic case.

% summary
In this paper, we study ``single-site'' genetic statistics,
i.e., statistics of aligned genome sequence that can be expressed as averages over values computed
separately for each site. We develop a theoretical and computational framework
that encompasses a large class of population genetic statistics, generalising
many classical summaries of genetic variation. We define
and explore a duality between ``Site'' and ``Branch'' statistics, and
provide a detailed description (and analysis) of an efficient algorithm
to compute statistics in this general setting.
The methods we describe are implemented, tested, and validated in the \tskit{}
Python and C libraries, freely available at \url{https://github.com/tskit-dev/tskit}
under the terms of the MIT license.

@@

% visualization and time decomposition adding a new dimension
Genomic data are naturally distributed on two axes: along the genome, and across geography.
The tree sequence extends this to a third axis: time.
A great many methods in population genetics aim to describe aspects of history,
and accurate (or at least unbiased) tree sequence inference
may shift the focus of these problems from inference to descriptive analysis.
The methods developed here distribute the contribution to various statistics across each tree,
and so could also be used to summarize contributions to various statistics across time.
This could provide, for instance, the time distribution of mutations or branches
contributing positively or negatively to $f_4$ statistics of introgression,
enabling historical interpretations of these signals.
% This additional mode of reporting statistics is an important subject for future work.
% moving past a few summaries in discrete groups to a genome/time/space view of relatedness
The computational toolbox of population genetics
is still mostly composed of statistics orginally designed for analysis of a handful of loci
in a small number of discrete, mostly separated populations.
Both our data and our understanding of the world have moved beyond this.
We hope that the tools developed here will help make it possible
to visualize and analyze genetic variation and genealogical relatedness
along the genome, across space, and through time.

>>>>

Genotype matrices are massively redundant, however, and several specialized
compression methods have been proposed~\citep{
christley2008human,qiao2012handling,durbin2014efficient,sambo2014compression,layer2016efficient,
danek2018gtc,lin2019sparse}.
The fundamental reason for this redundancy is that samples share ancestry:
related individuals will tend to share the same state at a particular variant site.
Trees describe these ancestral
relationships~\citep{felsenstein2004inferring,semple2003phylogenetics},
and provide a natural and elegant way of encoding genetic data,
not only recording the history of a sample but also massively compressing
the genotype matrix~\citep{ane2005missing}. Until recently, however,
tree based approaches could not be used to encode data from sexually
reproducing organisms as recombination results in many trees along the
genome~\citep{griffiths1991two,griffiths1996ancestral,rasmussen2014genome},
and there was no efficient way of representing this information.
The \emph{succinct tree sequence} (or \emph{tree sequence}, for brevity)
is a recently introduced data structure that
encodes these genealogical trees along a genome concisely.
It was introduced in the context of coalescent
simulation~\citep{kelleher2016efficient}, leading to scalability increases of
several orders of magnitude over existing methods. The methods were subsequently extended
and refined for forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree}, with similarly large efficiency gains.
Recent work has shown that tree sequence algorithms can
also be used to massively increase the scalability of methods for inferring
genome-wide genealogies, and making it possible to infer trees for millions of
samples~\citep{kelleher2019inferring}.
The key to the remarkable efficiency of tree sequence
algorithms is the way that shared structure in adjacent trees along the genome is encoded.
As one looks across the genome, genealogies change at recombination breakpoints in ancestors.
However, nearby trees tend to share much of their structure,
and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
are often shared across relatively long distances,
which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
This redundancy can be exploited to store genome sequence and the
associated genealogical relationships very compactly, and has formed the basis for
efficient algorithms to calculate several population genetics
statistics~\citep{kelleher2016efficient}.
This paper generalizes the approach to a much broader class of statistics,
supporting arbitrary tree topologies and patterns of allelic variation.
% Want to emphasise generality here: people might assume we're only dealing
% with the biallelic case.

% summary
In this paper, we study ``single-site'' genetic statistics,
i.e., statistics of aligned genome sequence that can be expressed as averages over values computed
separately for each site. We develop a theoretical and computational framework
that encompasses a large class of population genetic statistics, generalizing
many classical summaries of genetic variation. We define
and explore a duality between ``Site'' and ``Branch'' statistics, and
provide a detailed description (and analysis) of an efficient algorithm
to compute statistics in this general setting.
The methods we describe are implemented, tested, and validated in the \tskit{}
Python and C libraries, freely available at \url{https://github.com/tskit-dev/tskit}
under the terms of the MIT license.

@@

% visualization and time decomposition adding a new dimension
Genomic data are naturally distributed on two axes: along the genome, and across geography.
The tree sequence extends this to a third axis: time.
A great many methods in population genetics aim to describe aspects of history,
and accurate (or at least unbiased) tree sequence inference
may shift the focus of these problems from inference to descriptive analysis.
The methods developed here distribute the contribution to various statistics across each tree,
and so could also be used to summarize contributions to various statistics across time.
This could provide, for instance, the time distribution of mutations or branches
contributing positively or negatively to $f_4$ statistics of introgression,
enabling historical interpretations of these signals.
% This additional mode of reporting statistics is an important subject for future work.
% moving past a few summaries in discrete groups to a genome/time/space view of relatedness
The computational toolbox of population genetics
is still mostly composed of statistics originally designed for analysis of a handful of loci
in a small number of discrete, mostly separated populations.
Both our data and our understanding of the world have moved beyond this.
We hope that the tools developed here will help make it possible
to visualize and analyze genetic variation and genealogical relatedness
along the genome, across space, and through time.
